---
title: "Weather and Corn Yield Regressions"
author: "Casey Menick"
date: "2/25/2022"
output: 
  html_document:
    toc: true
    toc_float: false
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
library(ggiraphExtra)
```

# Iowa Weather Data Analysis

## Load PRISM daily maximum temperatures for Iowa

```{r tmax data, warning = FALSE, message = FALSE, results="hide"}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (째C)") +
  ggtitle("Daily Maximum Temperature in Iowa County Code #1")


```

```{r tidying up, warning = FALSE, message = FALSE, results="hide"}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Exploring Data in Winneshiek County, Iowa

### Summer temperature trends: Linear Regression

```{r temp trends, warning = FALSE, message = FALSE}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "temperature (째C)", title="Maximum Average Summer Temperature Trends in Winneshiek, Iowa", subtitle = "Linear regression line with year as predictor for maximum summer temperature") +
  geom_smooth(method = "lm", se=FALSE)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter temperature trends: Linear Regression

```{r winter temps, warning = FALSE, message = FALSE}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "temperature (째C)", title="Maximum Average Winter Temperature Trends in Winneshiek, Iowa", subtitle = "Linear regression line with year as predictor for maximum winter temperature") +
  geom_smooth(method = "lm", se=FALSE)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Winter temperature trends: Quadratic Regression

```{r quadratic temp trend, warning = FALSE, message = FALSE}

winnewinter$yearsq <- winnewinter$year^2
lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted),color="blue") +
  theme_bw() +
  labs(x = "year", y = "temperature (째C)", title="Maximum Average Winter Temperature Trends in Winneshiek, Iowa", subtitle = "Quadratic regression line with year and year^2 as predictors") 

summary(lm_wintertmaxquad)

```
# Iowa Crop Yield Analysis

## Download NASS corn yield data

```{r yield download, warning = FALSE, message = FALSE, results='hide'}

# set our API key with NASS
nassqs_auth(key = "C81C90AE-5B51-325E-8F57-FAF3E0173E0E")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```


## Exploring Data in Winneshiek County, Iowa

### Corn Yield Trends: Linear Regression with Time

Corn yields in Iowa have been increasing at a linear rate over time since 1980

```{r, warning = FALSE, message = FALSE}
winneshiek <- cornyields %>% 
  filter(county_name == "WINNESHIEK")

lm_winne1 <- lm(yield~year,winneshiek)
winneshiek$lm1fit <- lm_winne1$fitted.values

summary(lm_winne1)

ggplot(winneshiek,aes(year,yield)) +
  geom_point()+
  geom_line(aes(year,lm1fit),color="blue") +
  labs(title="Winneshiek County, Iowa Corn Yields Over Time", subtitle = "Linear model with year as predictor for yield") +
  theme_bw()
```
### Corn Yield Trends: Quadratic Regression with Time

Adding in a year-squared term to our linear time model does not improve our estimations.

```{r, warning = FALSE, message = FALSE}
winneshiek$yearsq <- winneshiek$year^2

lm_winne2 <- lm(yield ~ year + yearsq, winneshiek)
winneshiek$lm2fit <- lm_winne2$fitted.values

summary(lm_winne2)

ggplot(winneshiek,aes(year,yield)) +
  geom_point() +
  geom_line(mapping = aes(x = year, y = lm2fit)) +
  theme_bw() +
  labs(x = "year", y = "yield") +
  labs(title="Winneshiek County, Iowa Corn Yields Over Time", subtitle = "Linear model with year and year^2 as predictors for yield")
```

### Corn Yield Trends: Linar and Quadratic Regression with Temperature
When comparing the relationship between temperature and yield for Winneshiek, several models were compared. A linear regression of max temp, a linear regression of max temp and year, and a quadratic regression of max temp and max temp squared were assessed. Our second linear model using both max temp and year had the highest performance.

```{r, warning = FALSE, message = FALSE}
winneshiektemps <- left_join(winneshiek,winnesummer,by="year") %>% 
  mutate(meantmax2 = meantmax^2) %>% 
  drop_na()

lm_winne3 <- lm(yield ~ meantmax, winneshiektemps)
winneshiektemps$lm3fit <- lm_winne3$fitted.values
summary(lm_winne3)
ggplot(winneshiektemps,aes(meantmax,yield)) +
  geom_point() +
  geom_line(mapping = aes(x = meantmax, y = lm3fit)) +
  theme_bw() +
  labs(title="Winneshiek County, Iowa Corn Yields in Relation to Average Max Summer Temp", subtitle = "Linear model with year as predictor for yield")

lm_winne31 <- lm(yield ~ meantmax + year , winneshiektemps)
winneshiektemps$lm31fit <- lm_winne31$fitted.values
summary(lm_winne31)
ggPredict(lm_winne31) +
  theme_bw() +
  labs(title="Winneshiek County, Iowa Corn Yields in Relation to Average Max Summer Temp", subtitle = "Linear model with max temp and year as predictors for yield")

lm_winne32 <- lm(yield ~ meantmax + meantmax2, winneshiektemps)
winneshiektemps$lm32fit <- lm_winne32$fitted.values
summary(lm_winne32)
ggplot(winneshiektemps,aes(meantmax,yield)) +
  geom_point() +
  geom_line(mapping = aes(x = meantmax, y = lm32fit)) +
  theme_bw() +
  labs(title="Winneshiek County, Iowa Corn Yields in Relation to Average Max Summer Temp", subtitle = "Linear model with max temp and max temp^2 as predictors for yield")
```

## Exploring Data Across Counties

### Corn Yields: Linear Regression with Temperature, 2018

```{r, warning = FALSE, message = FALSE}
corn2018 <- left_join(cornyields %>% 
                        filter(year ==2018)%>% 
                        select(-year) %>% 
                        mutate(county_ansi = as.factor(county_ansi)),
                      tmaxdf %>%
                        filter(doy >= 152 & doy <= 243, year==2018) %>%
                        group_by(year,countyfp) %>%
                        summarize(meantmax = mean(tmax)),
                      by=c("county_ansi"="countyfp"))

lm_corn18 <- lm(yield~meantmax,corn2018)
corn2018$lmfit <- lm_corn18$fitted.values
summary(lm_corn18)

ggplot(corn2018,aes(meantmax,yield)) +
  geom_line(mapping = aes(x = meantmax, y = lmfit)) +
  geom_point() +
  labs(title="Iowa Corn Yields for all Counties in Relation to 2018 Average Max Summer Temp", subtitle = "Linear model with max temp as predictor for yield") +
  theme_bw()
```

### Corn Yield Trends: Panel Regression with county, year, max temp, max temp squared

```{r, warning = FALSE, message = FALSE}
allsummer <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(year,countyfp) %>%
  summarize(meantmax = mean(tmax))

cornall <- left_join(cornyields %>% 
                        mutate(county_ansi = as.factor(county_ansi)),
                      tmaxdf %>%
                        filter(doy >= 152 & doy <= 243) %>%
                        group_by(year,countyfp) %>%
                        summarize(meantmax = mean(tmax)),
                      by=c("county_ansi"="countyfp","year"="year"))%>% 
  drop_na() %>% 
  mutate(meantmax2=meantmax^2)

lm_cornpanel <-lm(yield~county_ansi+meantmax+meantmax2+year,cornall)
cornall$lmfit <- lm_cornpanel$fitted.values

ggplot(cornall,aes(meantmax,yield)) +
  geom_point(alpha=0.2) +
  geom_smooth(se=FALSE,method="lm",color="black") +
  theme_bw() +
  labs(x="Maximum Temperature (C)", title="Iowa Corn Yields for all Counties in Relation to Average Max Summer Temp", subtitle="Linear regression of maximum summer temperature in relation to yield")

summary(lm_cornpanel)
ggplot(cornall,aes(yield,lmfit))+
  geom_point(alpha=0.2) +
  geom_abline(size=1) +
  labs(title="Comparing Regression Fitted Values to Actual Yields for Iowa County Corn Yields", subtitle = "Panel regression of max temp, max temp^2,year, and county in relation to yield") +
  theme_bw() +
  xlab("Actual Corn Yield") +
  ylab("Predicted Corn Yield")
```

### Comparing Corn and Soybean Yields: Comparing growth over time, effect of maximum temperature, and effectiveness of the same panel regression

```{r, results='hide', message=FALSE,warning=FALSE}
# set our API key with NASS
nassqs_auth(key = "C81C90AE-5B51-325E-8F57-FAF3E0173E0E")

# parameters to query on 
paramssoy <- list(commodity_desc = "SOYBEANS", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soyyieldsall <- nassqs_yields(paramssoy)

soyyieldsall$county_ansi <- as.numeric(soyyieldsall$county_ansi)
soyyieldsall$Value <- as.numeric(soyyieldsall$Value)

# clean and filter this dataset
soyyields <- soyyieldsall %>% 
  select(county_ansi, county_name, Value, year) %>%
  filter(county_ansi>0 & !is.na(Value))
soyyields <- tibble(soyyields) %>% 
  mutate(Value = as.numeric(Value),
         county_ansi=as.factor(county_ansi))

# compare corn and soy yields
allia <- left_join(cornall,soyyields,by=c("year"="year","county_ansi" = "county_ansi")) %>% 
  drop_na() %>% 
  group_by(year) %>% 
  summarise(corn=mean(yield),
            soy=mean(Value)) %>% 
  pivot_longer(.,c("corn","soy"),names_to="crop")

ggplot(allia,aes(year,value,color=crop)) +
  geom_point() +
  geom_smooth(method="lm",se=FALSE) +
  theme_bw() +
  labs(x="yield", title = "Comparing County Average for Corn and Soy Yields Over Time", subtitle="Linear model with year as predictor for yield")

#panel
allia2 <- left_join(cornall,soyyields,by=c("year"="year","county_ansi" = "county_ansi")) %>% 
  drop_na() %>% 
  group_by(year,county_ansi) %>% 
  summarise(corn=mean(yield),
            soy=mean(Value),
            meantmax=mean(meantmax),
            meantmax2=meantmax^2)

ggplot(allia2,aes(meantmax,soy)) +
  geom_point(alpha=0.2) +
  geom_smooth(se=FALSE,method="lm",color="black") +
  theme_bw()+ 
  labs(x="Maximum Temperature (C)", title = "Comparing County Average for Corn and Soy Yields in Relation to Maximum Summer Temperature", subtitle="Linear model with maximum temperature as predictor for yield")


ggplot(allia2 %>% pivot_longer(.,c("corn","soy"),names_to="crop"),aes(meantmax,value,color=crop)) +
  geom_point(alpha=0.2) +
  geom_smooth(method="lm",se=FALSE) +
  theme_bw() +
  labs(x="Maximum Temperature (C)", title="Comparing Regression Fitted Values to Actual Yields for Iowa County Corn Yields", subtitle = "Panel regression of max temp, max temp^2,year, and county in relation to yield")

lm_soypanel <-lm(soy~county_ansi+meantmax+meantmax2+year,allia2)
allia2$lmfit <- lm_soypanel$fitted.values

summary(lm_soypanel)
ggplot(allia2,aes(soy,lmfit))+
  geom_point(alpha=0.2) +
  geom_abline(size=1) +
  labs(title="Comparing Regression Fitted Values to Actual Yields for Iowa County Soy Yields", subtitle = "Panel regression of max temp, max temp^2,year, and county in relation to yield") +
  theme_bw() +
  xlab("Actual Soy Yield") +
  ylab("Predicted Soy Yield")
```

Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.
